Parallel global optimization on GPU \thanks{The research is supported by the grant of the Ministry of education and science of the Russian Federation (the agreement of August 27, 2013 No 02.B.49.21.0003)}

### 2019-10-07 2016/JoGO_GPU/BarkalovGergel.tex
JOGO2016 paper next version

This present paper considers parallel algorithms for solving multiextremal optimization problems. The objective function in such problems is characterized by existence of several local optima (usually their number is not known and can be large). In multiextremal problems the possibility of a reliable estimation of the global optimum is essentially based on the availability of an a priori information about the function, allowing us to link possible values of an optimizable function with known values at search points. Very often, such information about the problem is presented as the assumption that the objective function $\varphi$ satisfies a Lipschitz condition with an a priori unknown constant $L$. The objective function can be represented by the algorithm for calculating its values in the search domain, i.e. it can be a ``black-box'' function. Similar problems are common in applications (problems of optimal design of objects and technological processes in various fields of technology, problems of model fitting according to observed data in scientific research, optimization of control processes, etc., see the review \cite{RefPinter}).

@@

One of the promising trends in the field of parallel global optimization (as in many areas connected with software implementation of complex algorithms) is the use of graphics processing units (GPUs). During the last decade, productivity of GPUs has been rapidly growing to satisfy the continuously growing demand of graphic application developers. Furthermore, some principles of development of graphics hardware have changed over the past few years. Now graphics hardware is more programmable. Today a GPU is a high-performance flexibly programmable and massively parallel processor that can provide solution to many computationally complex problems \cite{RefHwu}.

However, in the field of global optimization the potential of GPUs is not completely utilized. GPUs are generally used for parallelization of algorithms, which one way or another are based on the concept of random search (see works \cite{RefFerreiro,RefZhu,RefGarcia,RefMussi}); a review of this trend of GPU optimization algorithms is given in \cite{RefLangdon}. Owing to their stochastic nature, algorithms of this type generally do not guarantee convergence to a globally optimal solution, and that sets them apart from deterministic methods.

Some parallel variants are proposed for many deterministic algorithms of lipschitzian global optimization with guaranteed convergence \cite{RefGergel2005,RefEvtushenko,RefHe,RefPaulavicius}. However, these versions of algorithms are parallelized on CPU using MPI and/or OpenMP technologies; there are no implementations on GPU. The present article contains the results of an investigation into the GPU implementation of the global search parallel algorithm developed according to the information-statistical approach presented in the monograph \cite{RefStrongin2000}.

@@

There is a number of ways to adapt effective one-dimensional algorithms for solving multidimensional problems; see, for example, the diagonal partitions method in \cite{RefSergeyev2006} or the simplicial partitions method in \cite{RefZilinskas}. In this paper, we will use the approach based on the idea of dimension reduction by means of a Peano curve $y(x)$, which continuously and unambiguously maps the unit interval [0,1] onto the $n$-dimensional cube
\[
\left\{y\in R^N: -2^{-1}\leq y_i \leq 2^{-1}, 1 \leq i \leq N\right\}=\left\{y(x):0\leq x \leq 1 \right\}.
\]
Problems of numerical construction of Peano-type space filling curves and the corresponding theory are considered in detail in \cite{RefStrongin2000}, \cite{RefSergeyev2013}. Here we will note that a numerically constructed curve (\textit{evolvent}) is an approximation to a theoretical Peano curve with an accuracy $2^{-m}$ , where $m$ is an evolvent construction parameter. It means that using an evolvent we can distinguish points in a multidimensional space, the coordinates of which differ by not less than $2^{-m}$; closer points are considered as coincident. Thereby an evolvent constructed with a fixed m allows obtaining a solution of a problem with an accuracy of not more than $2^{-m}$ by coordinate.

@@

Therefore, it is possible, without limitation of generality, to consider minimization of one-dimensional function
\[
f(x)=\varphi(y(x)), \;\; x\in[0,1],
\]
satisfying the H{\"o}lder condition.

@@

The algorithm terminates if the condition $\Delta_t<\epsilon$ is satisfied; here $\epsilon>0$ the accuracy is preset. For estimation of the global solution, values
\[
f_k^\ast=\min_{1\leq i \leq k}f(x^i), \ x_k^\ast=\arg \min_{1\leq i \leq k}f(x^i).
\]
are selected.

Remark 1. The normalized characteristics $R(i)$ from (\ref{eq:14}) can be viewed as probabilities of locating a global minimum within the interval $[x_{i-1},x_i], 1\leq i \leq k+1$. This statement is proved in \cite{RefStrongin2000}. Thus, at each iteration a new trial point is selected inside the interval, which has the greatest probability of finding a global minimum.

@@

Let us illustrate the work of this \textit{global search algorithm} (GSA) using a dimension reduction for minimization of a two-dimensional multiextremal function
\begin{eqnarray} \nonumber \label{eq:19}
\varphi(y)= -&\left\{\left(\sum^{7}_{i=1}\sum^{7}_{j=1}A_{ij}g_{ij}(y)+B_{ij}h_{ij}(y)\right)^2+\right. \\
&\left.\left(\sum^{7}_{i=1}\sum^{7}_{j=1}C_{ij}g_{ij}(y)+D_{ij}h_{ij}(y)\right)^2\right\}^{1/2},\\ \nonumber
\end{eqnarray}
where
\begin{eqnarray} \nonumber
& y=(y_1,y_2)\in R^2, 0 \leq y_s \leq 1, s=1,2, \\ \nonumber
& g_{ij}(y)=\sin(i\pi y_1)\sin(j\pi y_2),  \\ \nonumber
& h_{ij}(y)=\cos(i\pi y_1)\cos(j\pi y_2), \nonumber 
\end{eqnarray}
and coefficients $A_{ij}, B_{ij}, C_{ij}, D_{ij}$  are taken uniformly in the interval $[-1,1]$. The experiment used the method parameters $r=3$, $\epsilon=10^{-3}$, and the evolvent construction parameter $m=12$. Fig.~\ref{fig:1} shows the contour lines of a function generated according to (\ref{eq:19}) and the points of 1103 trials carried out using the method before the required accuracy was obtained.
\begin{figure}
	\center
  \includegraphics[width=0.75\textwidth]{fig1.jpg} 
  \caption{Minimization of a two-dimensional function by the sequential algorithm}
  \label{fig:1}       % Give a unique label
\end{figure}

@@

The first option is to parallelize calculation of the target function describing the optimized object. This method can provide some speedup, but is specific for each problem.

@@

The third option is to  change the scheme of an algorithm for the purpose of carrying out several trials in parallel. This approach is most promising since it is characterized by efficiency (the part of computing process, in which the bulk of calculations is carried out, is parallelized) and generality (it is applicable for a wide class of characteristical algorithms of multiextremal optimization \cite{RefGrishagin1997}).

@@

Steps 1--3 of the parallel algorithm fully repeat steps 1--3 of the global search algorithm.

@@

Let us use a parallel algorithm of global search for solving a problem from section \ref{sec:2} with the numbers of parallel trials $p=2$ and $p=4$. With $p=2$ the number of iterations was $569$, and the number of trials was $1138$. With $p=4$ the number of iterations was reduced to $270$, and $1080$ trials were carried out. Fig.~\ref{fig:2} shows contour lines of the function with points of the trials performed by the method with $p=2$. Points of redundant trials carried out using a parallel algorithm are marked in the figure by "+". The concept of redundancy is discussed in more detail in the following section.
\begin{figure}
	\center
  \includegraphics[width=0.75\textwidth]{fig2.jpg} 
  \caption{Minimization of a two-dimensional function by the parallel algorithm}
  \label{fig:2}
\end{figure}

@@

What are the advantages of GPU over CPU when implementing global optimization algorithms? First, CPU has only a small (up to $16$ in the advanced models) number of cores working independently at a high clock frequency. Each CPU core is a powerful computer. GPU, on the contrary, works at low clock frequencies and uses hundreds of simpler computing elements. Second, a considerable part of a CPU crystal is occupied by a high-speed cache memory, while practically the whole of GPU consists of arithmetic and logic units. Therefore, GPUs are especially effective in problems, in which the number of arithmetic operations is great in comparison with memory operations.

With respect to methods of global optimization, parallel computing of many values of the objective function simultaneously is an operation which can be implemented effectively on GPU. In the function value calculation procedure implemented on GPU transfers of information from CPU to GPU will be minimal: it is necessary only to transfer the coordinates of trial points to GPU and to receive the values of the function in these points. The functions that determine the processing of trial results according to algorithm and require working with a large volume of collected search information can be effectively implemented on CPU.

@@

\begin{figure}
	\center
  \includegraphics[width=0.75\textwidth]{fig3.jpg} 
  \caption{Scheme of information exchanges in a GPU algorithm}
  \label{fig:3}
\end{figure}

@@

It is noteworthy that an important parameter for starting any GPU program (unlike a parallel program on CPU) is not only the number of threads $p$, but also the size of the \textit{thread block}. Separate threads on GPU are grouped in blocks of threads of identical size, thus each thread block is executed on a separate multiprocessor. Threads in a thread block can interact effectively using information in shared memory and synchronization. For the purpose of assessing the importance of the size of a thread block during the execution period of a GPU algorithm let us perform a preliminary series of experiments with the test function (\ref{eq:33}) with $N=5$. Let us vary the number $p$ of the used GPU threads and the size $pB$ of a thread block. All the other parameters of the method do not change ($r=2.3$, $\epsilon = 10^{-3}$, the evolvent construction parameter $m=10$). Table \ref{tab:2} shows the GPU algorithm operating time depending on $p$ and $pB$.
\begin{table}
	\caption{GPU algorithm operating time depending on thread block size}
	\label{tab:2}
	\center
	\begin{tabular}{llllll}
		\hline\noalign{\smallskip}
%		 $p$ & $pB=16$ & $pB=32$ & $pB=64$ & $pB=128$ & $pB=256$  \\
		$p$ & \multicolumn{5}{l}{ \hfil $pB$ \hfil } \\
		\noalign{\smallskip} \cline{2-6} \noalign{\smallskip}
		 & 16 & 32 & 64 & 128 & 256  \\
		\noalign{\smallskip} \hline \noalign{\smallskip}
		100 &	39.43 &	41.5 &	39.71 &	39.85 &	39.61 \\
		300 &	24.84 &	24.52 &	24.54 &	24.66 &	25.08 \\
		500 &	9.25 &	9.45 &	9.24 &	9.4 &	9.3\\
		700 &	11.8 &	13.43 &	11.78 &	11.91 &	12.03\\
		900 &	10.9 &	10.91 &	10.85 &	10.86 &	10.96\\
		1000 &	10.92 &	10.8 &	10.76 &	10.9 &	10.74\\
		\noalign{\smallskip}\hline
	\end{tabular}
\end{table}

As is seen from this table, there is a speedup of the algorithm in the case of an increase in the number of trials at one iteration up to the number of cores on GPU, with subsequent saturation. Owing to a weak information dependence the size of the thread block $pB$ does not have decisive effect, therefore let us use $pB=64$ in all subsequent experiments. When carrying out this set of experiments the load on GPU was 15 -- 85 \% (depending on the number of threads); the load was measured by GPUZ program.

After the parameters of the GPU algorithm have been determined let us make the main computing experiments on a series of multidimensional multiextremal problems. In work \cite{RefGaviano} a GKLS generator allowing generation of multiextremal optimization problems with known properties (number of local minima, size of their domains of attraction, global minimum point, function value, etc.) is described.

The results of numerical comparison of the three sequential algorithms -- DIRECT \cite{RefJones}, DIRECT\textit{l} \cite{RefGablonsky} and global search algorithm (GSA) from section \ref{sec:2} -- are provided below (results of the work of the first two algorithms are given in paper \cite{RefGaviano}). Numerical comparison was carried out on function classes \textit{Simple} and \textit{Hard} of dimension 4 and 5 from \cite{RefGaviano} since the solution of problems of dimension 2 and 3 demands a small number of iterations and it is impractical to use GPU for solving these problems. Global minimum $y^\ast$ was considered to be found, if the algorithm generated trial point $y^k$ in $\delta$-vicinity of the global minimum, i.e. $\left\|y^k-y^\ast\right\|\leq\delta$. The size of the vicinity was selected (according to \cite{RefGaviano}) as $\delta = \left\|b-a\right\|\sqrt[N]{\Delta}$, $N$ -- problem dimension, $a$ and $b$ -- borders of search domain $D$, parameter $\Delta=10^{-6}$ at $N=4$ and $\Delta=10^{-7}$ at $N=5$. When using the GSA for the class \textit{Simple} parameter $r=4.5$ was selected, for the class \textit{Hard} $r=5.6$ was selected; Peano curve construction parameter was fixed as $m=10$. The maximum allowable number of iterations was $K_{max} = 10^6$.

The average number of iterations $k_{av}$ performed by the method for solving a series of problems from these classes is shown in table \ref{tab:3}. Symbol $>$ reflects a situation in which not all problems of the class were solved by the method. This means that the algorithm was stopped as the maximum allowable number of iterations $K_{max}$ was achieved. In this case, value $K_{max}=10^6$ was used for calculating the average value of the number of iterations $k_{av}$ that corresponds to the lower estimate of this average value. The number of unsolved problems is specified in brackets.

\begin{table}
	\caption{Average number of iterations $k_{av}$}
	\label{tab:3}
	\center
	\begin{tabular}{lllll}
		\hline\noalign{\smallskip}
  	 $N$ & Problem class & DIRECT & DIRECT\textit{l} & GSA \\
		\noalign{\smallskip} \hline \noalign{\smallskip}
			4 &	\textit{Simple}	& $>$47282(4) &	18983 &	11953 \\
	      & \textit{Hard} &	$>$95708(7) &	68754 &	25263 \\
			5	& \textit{Simple} &	$>$16057(1) &	16758 &	15920 \\
				& \textit{Hard} &	$>$217215(16) &	$>$269064(4) & $>$148342 (4) \\
		\noalign{\smallskip}\hline
	\end{tabular}
\end{table}

@@

Let us make a similar experiment using a GPU version of the algorithm. Also let us fix the average number of iterations and average time spent on one problem (see tables \ref{tab:last4} -- \ref{tab:last}). Let us measure the speedup and redundancy of GPU algorithm in relation to the CPU algorithm started with 32 cores (see column $p=32$ in tables \ref{tab:4} and \ref{tab:5}). The number of the used threads $p$ on GPU varied, all other parameters of the method did not change. The number of threads, which is greater, than the number of cores, was chosen to show the following effect: the number of iterations executed by the algorithm will be reduced in case of an increase in the number of threads, while the algorithm operation time will not be reduced. The minimum time of operation is reached, if the number of used threads does not surpass the number of GPU cores.

@@

By now, parallel implementations have been suggested for the most of the existing sequential algorithms. A review of modern methods of parallel global optimization can be found, for example, in \cite{RefApuzzo}.

The methods described in \cite{RefApuzzo} can be divided into two classes: deterministic methods of global optimization (including the algorithm of global search considered in this article), and heuristic methods, which are one way or another based on the random search concept. Algorithms belonging to the latter class allow use of massive parallelism and are easily implemented on GPU (see works \cite{RefFerreiro,RefZhu,RefGarcia,RefMussi} and the review \cite{RefLangdon}).

Global optimization of implementations on GPU for deterministic algorithms has not been suggested yet. For example, parallelization of algorithms based on the ideas of the branch and bound method is carried out for a small number of processors using OpenMP and MPI \cite{RefPaulavicius}. It is noteworthy that the branch and bound method is applied in other areas of optimization; in particular, combinatorial optimization and integer programming. Using GPU shows good results in these areas (see \cite{RefBoukedjar}, \cite{RefCarneiro}). It is explained by the following factor. Problems of this class (such as the Travelling Salesman Problem or the Knapsack problem) are characterized by short time of criterion value calculation in one point and, at the same time, relatively long time required to process the results. Therefore, the most labor-intensive operations here are branching and bound computation, which are implemented on GPU (see \cite{RefBoukedjar}, \cite{RefCarneiro}).

This paper considers the problems of lipschitzian global optimization that are frequently encountered in applications. They are characterized by a considerable time required to calculate the function value in one point (in comparison with the processing of the results of such calculation). Therefore, calculation of the value of an optimized function in one point is done using a GPU, and the main objective of the algorithm (working at CPU) is to efficiently select points of parallel trials. This is carried out in the parallel implementation of GSA on GPU.

@@

@@

@@

\bibitem{RefApuzzo}
M. D’Apuzzo, M. Marino, A. Migdalas, P.M. Pardalos, and G. Toraldo, Parallel computing in global optimization, in Handbook of Parallel Computing and Statistics, E.J. Kontoghiorghes, ed. Chapman \& Hall, London, 2006, pp. 225–258

\bibitem{RefBoukedjar}
Boukedjar A., Lalami M. E., El Baz D. Parallel Branch and Bound on a CPU-GPU System //PDP. – 2012. – С. 392-398

\bibitem{RefCarneiro}
Carneiro T. et al. A new parallel schema for branch-and-bound algorithms using GPGPU //Computer Architecture and High Performance Computing (SBAC-PAD), 2011 23rd International Symposium on. – IEEE, 2011. – С. 41-47.

>>>>

This paper considers parallel algorithms for solving multiextremal optimization problems. The objective function in such problems is characterized by existence of several local optima (usually their number is not known and can be large). 
%new text begin
In multiextremal problems the possibility of a reliable estimation of the global optimum is essentially based on the availability of an a priori information about the function, allowing us to estimate lower bound of the objective function using known values at search points. 
%new text end
Very often, such information about the problem is presented as the assumption that the objective function $\varphi$ satisfies a Lipschitz condition with an a priori unknown constant $L$. 
%new text begin
The important feature of the problem under consideration is that the objective function is commonly not given by an analytical formula. The usual way to calculate the function value is to apply a computer-aided algorithm. Thus, each calculation may be interpreted as running some ``black-box'' to produce the outcome for the preset input. These runs could require substantial computer resources and therefore should not be too numerous. 
%new text end
%new word - such
Such problems are common in applications (problems of optimal design of objects and technological processes in various fields of technology, problems of model fitting according to observed data in scientific research, optimization of control processes, etc., see the review \cite{RefPinter}).

@@

One of the promising trends in the field of parallel global optimization (as in many areas connected with software implementation of complex algorithms) is the use of graphics processing units (GPUs). During the last decade, productivity of GPUs has been rapidly growing to satisfy the continuously growing demand of graphic application developers. Furthermore, some principles of development of graphics hardware have changed over the past few years. Today a GPU is a high-performance flexibly programmable and massively parallel processor that can provide solution to many computationally complex problems \cite{RefHwu}.

%new text begin
By now, parallel implementations have been suggested for most of the existing sequential algorithms. A review of modern methods of parallel global optimization can be found, for example, in \cite{RefApuzzo}. The methods described in \cite{RefApuzzo} can be divided into two classes: deterministic methods of global optimization, and heuristic methods, which are one way or another based on the random search concept. Algorithms belonging to the latter class allow the use of massive parallelism and are easily implemented on GPU (see works \cite{RefFerreiro,RefZhu,RefGarcia,RefMussi} and the review \cite{RefLangdon}). Owing to their stochastic nature, algorithms of this type generally do not guarantee convergence to a globally optimal solution, and that sets them apart from deterministic methods.

Some parallel variants are proposed for deterministic algorithms of lipschitzian global optimization with guaranteed convergence \cite{RefGergel2005,RefEvtushenko,RefHe,RefPaulavicius}. However, these versions of algorithms are parallelized on CPU; there are no implementations on GPU. For example, parallelization of algorithms based on the ideas of the branch and bound method is carried out for a small number of processors using OpenMP and MPI technologies \cite{RefPaulavicius}. It is noteworthy that the branch and bound method is applied in other areas of optimization; in particular, combinatorial optimization and integer programming. Using GPU shows good results in these areas \cite{RefBoukedjar}, \cite{RefCarneiro}. It is explained by the following factor. Problems of this class (such as the Travelling Salesman Problem or the Knapsack Problem) are characterized by short time of objective function value calculation in one point and, at the same time, relatively long time required to process the results. Therefore, the most labor-intensive operations here are branching and bound computation, which are implemented on GPU (see \cite{RefBoukedjar}, \cite{RefCarneiro}).

This paper considers the problems of lipschitzian global optimization that are frequently encountered in applications. They are characterized by a considerable time required to calculate the function value in one point (in comparison with the processing of the results of such calculation). Therefore, calculation of the value of an optimized function in one point can be done using a GPU, and the main objective of the algorithm (working at CPU) is to efficiently select points of parallel trials. The article contains the results of an investigation into the GPU implementation of the global search parallel algorithm developed according to the information-statistical approach presented in the monograph \cite{RefStrongin2000}.
%new text end

@@

There is a number of ways to adapt effective one-dimensional algorithms for solving multidimensional problems; see, for example, the diagonal partitions method in \cite{RefSergeyev2006} or the simplicial partitions method in \cite{RefZilinskas}. In this paper, we will use the approach based on the idea of dimension reduction by means of a Peano curve $y(x)$, which continuously and unambiguously maps the unit interval [0,1] onto the $n$-dimensional cube
\[
\left\{y\in R^N: -2^{-1}\leq y_i \leq 2^{-1}, 1 \leq i \leq N\right\}=\left\{y(x):0\leq x \leq 1 \right\}.
\]
%new text begin
Problems of numerical construction of Peano-type space filling curves and the corresponding theory are considered in detail in \cite{RefStrongin2000}, \cite{RefSergeyev2013}.  
%new text begin
Here we will note that a numerically constructed curve (\textit{evolvent}) is $2^{-m}$ accurate approximation of the theoretical Peano curve in $L_\infty$ metric, where $m$ is an evolvent construction parameter. Examples of the evolvent with different $m$ in two dimensions are given in Fig.~\ref{fig:0}.

%Here we will note that a numerically constructed curve (\textit{evolvent}) is an approximation to a theoretical Peano curve with an accuracy $2^{-m}$ , where $m$ is an evolvent construction parameter. It means that using an evolvent we can distinguish points in a multidimensional space, the coordinates of which differ by not less than $2^{-m}$; closer points are considered as coincident. Thereby an evolvent constructed with a fixed $m$ allows obtaining a solution of a problem with an accuracy of not more than $2^{-m}$ by coordinate.

\begin{figure}
\begin{minipage}{0.32\linewidth}
\center{\includegraphics[width=1.0\linewidth]{fig1a.JPG} \\ (a)}
\end{minipage}
\hfill
\begin{minipage}{0.32\linewidth}
\center{\includegraphics[width=1.0\linewidth]{fig1b.JPG} \\ (b)}
\end{minipage}
\begin{minipage}{0.32\linewidth}
\center{\includegraphics[width=1.0\linewidth]{fig1c.JPG} \\ (c)}
\end{minipage}
\caption{Elvolvents in two dimensions with (a) $m=3$, (b) $m=4$ and (c) $m=5$}
\label{fig:0}
\end{figure}
%new text end

@@

Therefore, it is possible, without limitation of generality, to consider minimization of one-dimensional function
\[
f(x)=\varphi(y(x)), \;\; x\in[0,1],
\]
satisfying the H{\"o}lder condition.
%new text begin
Once again, we recall that the Peano curve $y(x)$ is defined as a limit object emerging in some sequential construction. Appropriate approximations to $y(x)$ are to be used for practical application.
%new text end

@@

The algorithm terminates if the condition $\Delta_t<\epsilon$ is satisfied; here $\epsilon>0$ is the preset accuracy. For estimation of the global solution, values
\[
f_k^\ast=\min_{1\leq i \leq k}f(x^i), \ x_k^\ast=\arg \min_{1\leq i \leq k}f(x^i).
\]
are selected.

%new text begin
Remark 1. The global search algorithm was developed in the framework of a stochastic model (see \cite{RefStrongin2000}). From this point of view the normalized characteristics $R(i)$ from (\ref{eq:14}) can be considered as probabilities of locating the global minimum within the interval  $(x_{i-1},x_i), 1\leq i \leq k+1$. Thus, at each iteration a new trial point is selected inside the interval, which has the greatest probability of finding the global minimum. 
%new text end

@@

Let us illustrate the work of this \textit{global search algorithm} (GSA) using a dimension reduction for minimization of a two-dimensional multiextremal function
\begin{eqnarray} \nonumber \label{eq:19}
\varphi(y)= -&\left\{\left(\sum^{7}_{i=1}\sum^{7}_{j=1}A_{ij}g_{ij}(y)+B_{ij}h_{ij}(y)\right)^2+\right. \\
&\left.\left(\sum^{7}_{i=1}\sum^{7}_{j=1}C_{ij}g_{ij}(y)+D_{ij}h_{ij}(y)\right)^2\right\}^{1/2},\\ \nonumber
\end{eqnarray}
where
\begin{eqnarray} \nonumber
& y=(y_1,y_2)\in R^2, 0 \leq y_s \leq 1, s=1,2, \\ \nonumber
& g_{ij}(y)=\sin(i\pi y_1)\sin(j\pi y_2),  \\ \nonumber
& h_{ij}(y)=\cos(i\pi y_1)\cos(j\pi y_2), \nonumber 
\end{eqnarray}
and coefficients $A_{ij}, B_{ij}, C_{ij}, D_{ij}$  are taken uniformly in the interval $[-1,1]$. The experiment used the method parameters $r=3$, $\epsilon=10^{-3}$, and the evolvent construction parameter $m=12$. Fig.~\ref{fig:1} shows the contour lines of a function generated according to (\ref{eq:19}) and the points of 1103 trials carried out using the method before the required accuracy was obtained.
\begin{figure}
	\center
  \includegraphics[width=0.75\textwidth]{fig2.jpg} 
  \caption{Minimization of a two-dimensional function by the sequential algorithm}
  \label{fig:1}       % Give a unique label
\end{figure}

@@

The first option is to parallelize calculation of the objective function describing the optimized object. This method can provide some speedup, but is specific for each problem.

@@

%new text begin
The third option is to change an algorithm for the purpose of carrying out several trials in parallel. This approach is most promising since it is characterized by efficiency and generality. Firstly, the part of computing process, in which the bulk of calculations is carried out, is parallelized. Secondly, the approach is applicable for a wide class of characteristical algorithms of multiextremal optimization \cite{RefGrishagin1997}.
%new text end

@@

Steps 1--3 of the parallel algorithm fully repeat steps 1--3 of the sequential global search algorithm.

@@

Let us use a parallel algorithm of global search for solving a problem from section \ref{sec:2} with the numbers of parallel trials $p=2$ and $p=4$. With $p=2$ the number of iterations was $569$, and the number of trials was $1138$. With $p=4$ the number of iterations was reduced to $270$, and $1080$ trials were carried out. Fig.~\ref{fig:2} shows contour lines of the function with points of the trials performed by the method with $p=2$. Points of redundant trials carried out using a parallel algorithm are marked in the figure by "+". The concept of redundancy is discussed in more detail in the following section.
\begin{figure}
	\center
  \includegraphics[width=0.75\textwidth]{fig3.jpg} 
  \caption{Minimization of a two-dimensional function by the parallel algorithm}
  \label{fig:2}
\end{figure}

@@

What are the advantages of GPU over CPU when implementing global optimization algorithms? First, CPU has only a small (up to $16$ in the advanced models) number of cores working independently at a high clock frequency. GPU, on the contrary, works at low clock frequencies and uses hundreds of simpler computing elements. Second, a considerable part of a CPU crystal is occupied by a high-speed cache memory, while practically the whole of GPU consists of arithmetic and logic units. Therefore, GPUs are especially effective in problems, in which the number of arithmetic operations is great in comparison with memory operations.

With respect to methods of global optimization, parallel computing of many values of the objective function simultaneously is an operation which can be efficiently implemented on GPU. In the function value calculation procedure implemented on GPU transfers of information from CPU to GPU will be minimal: it is necessary only to transfer the coordinates of trial points to GPU and to receive the values of the function in these points. The functions that determine the processing of trial results according to algorithm and require working with a large volume of collected search information can be efficiently implemented on CPU.

@@

\begin{figure}
	\center
  \includegraphics[width=0.75\textwidth]{fig4.jpg} 
  \caption{Scheme of information exchanges in a GPU algorithm}
  \label{fig:3}
\end{figure}

@@

It is noteworthy that an important parameter for starting any GPU program (unlike a parallel program on CPU) is not only the number of threads $p$, but also the size of the \textit{thread block}. Separate threads on GPU are grouped in blocks of threads of identical size, thus each thread block is executed on a separate multiprocessor. Threads in a thread block can interact effectively using information in shared memory and synchronization. For the purpose of assessing the importance of the size of a thread block during the execution period of a GPU algorithm let us perform a preliminary series of experiments with the test function (\ref{eq:33}) with $N=5$. Let us vary the number $p$ of the used GPU threads and the size $p_b$ of a thread block. All the other parameters of the method do not change ($r=2.3$, $\epsilon = 10^{-3}$, the evolvent construction parameter $m=10$). Table \ref{tab:2} shows the GPU algorithm operating time depending on $p$ and $p_b$.
\begin{table}
	\caption{GPU algorithm operating time depending on thread block size}
	\label{tab:2}
	\center
	\begin{tabular}{llllll}
		\hline\noalign{\smallskip}
%		 $p$ & $p_b=16$ & $p_b=32$ & $p_b=64$ & $p_b=128$ & $p_b=256$  \\
		$p$ & \multicolumn{5}{l}{ \hfil $p_b$ \hfil } \\
		\noalign{\smallskip} \cline{2-6} \noalign{\smallskip}
		 & 16 & 32 & 64 & 128 & 256  \\
		\noalign{\smallskip} \hline \noalign{\smallskip}
		100 &	39.43 &	41.5 &	39.71 &	39.85 &	39.61 \\
		300 &	24.84 &	24.52 &	24.54 &	24.66 &	25.08 \\
		500 &	9.25 &	9.45 &	9.24 &	9.4 &	9.3\\
		700 &	11.8 &	13.43 &	11.78 &	11.91 &	12.03\\
		900 &	10.9 &	10.91 &	10.85 &	10.86 &	10.96\\
		1000 &	10.92 &	10.8 &	10.76 &	10.9 &	10.74\\
		\noalign{\smallskip}\hline
	\end{tabular}
\end{table}

As is seen from this table, there is a speedup of the algorithm in the case of an increase in the number of trials at one iteration up to the number of cores on GPU, with subsequent saturation. Owing to a weak information dependence the size of the thread block $p_b$ does not have decisive effect, therefore let us use $p_b=64$ in all subsequent experiments. When carrying out this set of experiments the load on GPU was 15 -- 85 \% (depending on the number of threads); the load was measured by GPUZ program.

After the parameters of the GPU algorithm have been determined let us make the main computing experiments on a series of multidimensional multiextremal problems. In work \cite{RefGaviano} a GKLS generator allowing generation of multiextremal optimization problems with known properties (number of local minima, size of their domains of attraction, global minimum point, etc.) is described.

The results of numerical comparison of the three sequential algorithms -- DIRECT \cite{RefJones}, DIRECT\textit{l} \cite{RefGablonsky} and global search algorithm (GSA) from section \ref{sec:2} -- are provided below (results of the work of the first two algorithms are given in paper \cite{RefGaviano}). Numerical comparison was carried out on function classes \textit{Simple} and \textit{Hard} of dimension 4 and 5 from \cite{RefGaviano} since solving a problems of dimension 2 and 3 demands a small number of iterations and it is impractical to use GPU for solving these problems. Global minimum $y^\ast$ was considered to be found, if the algorithm generated trial point $y^k$ in $\delta$-vicinity of the global minimum, i.e. $\left\|y^k-y^\ast\right\|\leq\delta$. The size of the vicinity was selected (according to \cite{RefGaviano}) as $\delta = \left\|b-a\right\|\sqrt[N]{\Delta}$, $N$ -- problem dimension, $a$ and $b$ -- borders of search domain $D$, parameter $\Delta=10^{-6}$ at $N=4$ and $\Delta=10^{-7}$ at $N=5$. When using the GSA for the class \textit{Simple} parameter $r=4.5$ was selected, for the class \textit{Hard} $r=5.6$ was selected; evolvent construction parameter was fixed as $m=10$. The maximum allowable number of iterations was $K_{max} = 10^6$.

The average number of iterations $k_{av}$ performed by the method for solving a series of problems from these classes is shown in table \ref{tab:3}. Symbol ''$>$''  reflects a situation in which not all problems of the class were solved by the method. This means that the algorithm was stopped as the maximum allowable number of iterations $K_{max}$ was achieved. In this case, value $K_{max}=10^6$ was used for calculating the average value of the number of iterations $k_{av}$ that corresponds to the lower estimate of this average value. The number of unsolved problems is specified in brackets.

\begin{table}
	\caption{Average number of iterations $k_{av}$}
	\label{tab:3}
	\center
	\begin{tabular}{lllll}
		\hline\noalign{\smallskip}
  	 $N$ & Problem class & DIRECT & DIRECT\textit{l} & GSA \\
		\noalign{\smallskip} \hline \noalign{\smallskip}
			4 &	\textit{Simple}	& $>$47282(4) &	18983 &	11953 \\
	      & \textit{Hard} &	$>$95708(7) &	68754 &	25263 \\
			5	& \textit{Simple} &	$>$16057(1) &	16758 &	15920 \\
				& \textit{Hard} &	$>$217215(16) &	$>$269064(4) & $>$148342(4) \\
		\noalign{\smallskip}\hline
	\end{tabular}
\end{table}

@@

Let us make a similar experiment using a GPU version of the algorithm. Also let us fix the average number of iterations and average time required to solve one problem (see tables \ref{tab:last4} -- \ref{tab:last}). Let us measure the speedup and redundancy of GPU algorithm in relation to the CPU algorithm started with 32 cores (see column $p=32$ in tables \ref{tab:4} and \ref{tab:5}). The number of the used threads $p$ on GPU varied, all other parameters of the method did not change. The number of threads, which is greater, than the number of cores, was chosen to show the following effect: the number of iterations executed by the algorithm will be reduced in case of an increase of the threads number, while the algorithm operation time will not be reduced. The minimum time of operation is reached, if the number of used threads does not surpass the number of GPU cores.

@@

@@

\bibitem{RefApuzzo}
D'Apuzzo, M.,  Marino, M., Migdalas, A., Pardalos, P.M., Toraldo, G.: Parallel computing in global optimization. In Handbook of Parallel Computing and Statistics, Chapman \& Hall, London, 225--258 (2006)

@@

\bibitem{RefBoukedjar}
Boukedjar A., Lalami M. E., El Baz D.: Parallel Branch and Bound on a CPU-GPU System. In 20th Int. Conf. on Parallel,  Distributed and Network-Based Processing, 392--398 (2012)

\bibitem{RefCarneiro}
Carneiro, T. et al.: A new parallel schema for branch-and-bound algorithms using GPGPU. In 23rd Int. Symposium on Computer Architecture and High Performance Computing, 41--47 (2011)

@@
